---
title: "Genomic Packages Tutorial"
author: "Miguel A. García-Campos"
output:
  html_document: 
    keep_md: yes
    theme: cerulean
    toc: yes
    toc_float: TRUE
    code_folding: show
---

```{r, Global Options, echo = F}
knitr::opts_chunk$set(warning = F, message = F)
```

# Motivation

These set of packages comprises many different parts of the process of genomic 
analysis. As these packages are all built with R, we can avoid the use of secondary
packages and programming environments. And as most of them are maintained through
the Bioconductor project, these packages are expected to work robustly and
updated in a coordinated manner.

# Setup

```{r, Setup}
# Cores to be used when parallel is used
nCores <- 10
```

## Packages

The packages we are going to introduce
```{r, packages and functions}
library(GenomicRanges)
library(GenomicAlignments)
library(Gviz)
library(magrittr)
library(BSgenome.Hsapiens.UCSC.hg19)
library(ggplot2)
```

## Functions

A group of custom functions
```{r, cache=T}
ggBarplot <- function(x, ci = NA, title = NULL, subt = NULL, xLab = "Names", yLab = "Values"){
    if(is.null(names(x))){names(x) <- 1:length(x)}
    df <- data.frame(names = factor(names(x), levels = names(x)), value = x, CI = ci)
    outGG <- ggplot(data=df, aes(x=names, y=value)) + 
        geom_bar(stat="identity") + theme_classic() + 
        geom_errorbar(aes(ymin=value-CI, ymax=value+CI), width=.2, position=position_dodge(.9)) + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle(title, subt) +
        ylab(yLab) + xlab(xLab)
    return(outGG)
}

```

# GenomicRanges

## Basic construction and usage

GenomicRanges are built over a package called IRanges. Implements an algebra of 
range operations, including efficient algorithms for finding overlaps and 
nearest neighbors. It is based on a base-1 (i.e. Genomes start at position 1)

The basic construction of a GRanges consist in the following code:

```{r, GRanges construction}
set.seed(30)
nReads <- 100
readSize <- 30
rStarts <- runif(n = nReads, min = 1, max = 200) %>% as.integer()
reads <- GRanges(seqnames = "chr1", 
                 ranges = IRanges(start = rStarts, end = rStarts + readSize),
                 strand = "+")
reads
```

## Essential functions to use with GRanges

`start()`: Extracts the start(s) of the Granges

`end()`: Extracts the end(s) of the Granges 

`strand()`: Extracts the strand of the Granges object

```{r, basic functions, cache = T}
start(reads) %>% head
end(reads) %>% head
strand(reads) %>% head
```

## findOverlaps()

`findOverlaps()` is perhaps the most important function of this package. In its 
most generic way, it allows to find which ranges overlap to other ranges. But its speed
and richness of information can accelerate many tasks for example counting the number of
reads that overlap an exon.

The output of the `findOverlaps()` function can be queried with the `@` operators
to extract the pairs of matches.

```{r, overlaps, cache=T}
(overlaps <- findOverlaps(query = reads[25], subject =  reads))
overlaps@from %>% head
overlaps@to %>% head
```

## coverage()

Another very useful function is the `coverage()` function. With these function
calculating the coverage of millions of reads is a task that takes a couple of seconds.

```{r, coverage}
coverage(reads)[["chr1"]] %>% as.numeric() %>% 
    barplot(main= "Coverage example", ylab = "Counts", xlab = "Position along chromosome")

```

Most elements inside a GRanges object get encoded into an RLE (Run Length Encoding) object.
For most of the manipulation of data RLE work as normal character or numeric vectors,
but in certain cases they have to be reverted to their "more conventional" version.

This can be done for example with the as.numeric() and as.character() functions.

```{r, showing RLE}
coverage(reads)[["chr1"]] %>% class
coverage(reads)[["chr1"]] 
coverage(reads)[["chr1"]] %>% as.numeric()
```

## Additional data in GRanges objects

Additional data can be added and retrieved using the "$" operator and the 
mcols() function
```{r, additional data table}
reads$readName <- paste0("read_", 1:nReads)
reads$passQC <- sample(x = c(T, F), prob = c(0.8, 0.2), size = nReads, replace = T)
mcols(reads)
```

## Importing BED files as range objects

For importing gene Annotations as BED or GFF files the rtracklayer package does 
this task very straighforwardly.

```{r, importing bed}
geneAnnot <- rtracklayer::import("/home/labs/schwartzlab/miguelg/BIGDATA/UCSC/hg19/geneAnnot_hg19_canonical.bed", format="bed")
geneAnnot
```

# Genomic Alignments

## Basic usage with paired-end libraries

Genomic alignments has been built as an interface to read alignment files into
R in an easy way.

We are going to load already aligned data using the `readGAlignmentPairs()` function
in the case of paired-end reads. (for single-end aligned reads we use the function
`readGAlignments()` function)

```{r, basic usage,  cache = T}
BAMfile <- "/home/labs/schwartzlab/miguelg/BIGDATA/GEO/GSE55572/bam/human_Exp2siControl_input_Aligned.out.sorted.bam"
genome <- BSgenome.Hsapiens.UCSC.hg19

flag0 <- scanBamFlag(isDuplicate = FALSE, # Base parameter
                     isNotPassingQualityControls=FALSE, 
                     isPaired = T) #Remove optical duplicates and low quality reads
iChr <- "chr13"
param0 <- ScanBamParam(flag=flag0, 
                       which = GRanges(seqnames= iChr, 
                                       ranges=IRanges(start=1, end= length(genome[[iChr]])),
                                       strand = "*"))
rds <- readGAlignmentPairs(BAMfile, use.names = TRUE, param = param0)
rds
```

## CIGAR and sequence information

To retrieve the CIGAR and sequence information, this has to be specified in the
scanning parameters using the `ScanBamParam()` function.

Additionally read information is stored separately in first and last reads. This
objects are accesible using the `@` operator.

```{r, basic usage II,  cache = T}
param1 <- ScanBamParam(flag=flag0, 
                       which = GRanges(seqnames= iChr, 
                                       ranges=IRanges(start=1, end= length(genome[[iChr]])),
                                       strand = "*"),
                       what = c("seq", "cigar"))
rds <- readGAlignmentPairs(BAMfile, use.names = TRUE, param = param1)
rds@first
```

For all chromosomes
```{r, basic usage III,  cache = T}
chroms <- seqnames(geneAnnot) %>% unique %>% as.character()
t0 <- Sys.time()
allReads <- lapply(chroms, function(iChr){
  param0 <- ScanBamParam(flag=flag0, 
                       which = GRanges(seqnames= iChr, 
                                       ranges=IRanges(start=1, end= length(genome[[iChr]])),
                                       strand = "*"))
  rds <- readGAlignmentPairs(BAMfile, use.names = TRUE, param = param0)
}) %>% do.call(what = c)
t1 <- Sys.time()
t1 - t0
length(allReads)
tmp <- seqnames(allReads) %>% table
tmp2 <- as.numeric(tmp); names(tmp2) <- names(tmp)
tmp2 <- sort(tmp2)
ggBarplot(tmp2) + theme_minimal() + ggtitle("Reads by chromosome") + 
  ylab("Reads") + xlab("Chromosomes") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# GViz

GViz is a package for visualization of genomic Data. Similar to IGV, but with 
more options and plotting capabilities.

## Demonstration

Using already analyzed locations we can visualize sections of a BAM file with
relative easyness. In this case we are plotting locations of putative SNPs 
detected in the human cell strain used in this study.

```{r, visualizing SNPs, cache = T}
load("/home/labs/schwartzlab/miguelg/WORKSPACE_wexac/m6A/allelicImbalance/allSNPs_ranges.Rdata")

axisTrack <- GenomeAxisTrack(exponent=0)
sTrack <- SequenceTrack(genome)
sTrack_neg <- SequenceTrack(genome, complement = T)
INPUTs_alTrack <- AlignmentsTrack("/home/labs/schwartzlab/miguelg/BIGDATA/GEO/GSE55572/bam/aggInput.bam", isPaired = T)
vidWindow <- 20

set.seed(seed = 745)
for(iSNP in sample(1:length(allSNPs_ranges), 5)){
  ideoTrack <- IdeogramTrack(genome = "hg19", chromosome = as.character(seqnames(allSNPs_ranges[iSNP])))
  plotTracks(list(ideoTrack, 
                  axisTrack,
                  INPUTs_alTrack,
                  sTrack,
                  sTrack_neg),
             chromosome = as.character(seqnames(allSNPs_ranges[iSNP])),
             from = start(allSNPs_ranges[iSNP]) - vidWindow,
             to = end(allSNPs_ranges[iSNP]) + vidWindow, min.height = 0,
             shape = "box", add53=TRUE, add35=TRUE, main = names(allSNPs_ranges[iSNP]))
}

```

# Session info

```{r, SessionInfo}
# Session Info
sessionInfo()
```
